### 2019-10-10 ch1_introduction/content.tex
update

\subsection{Cell dynamics and gene regulation}
Cells are dynamic entities that can gradually produce the molecules needed to acquire new functionality. The variation within and between cells happens because of differences in gene expression. The mechanism by which this happens is called gene regulation, and takes place primarily at the level of transcription. According to the needs of a cell, different genes are being transcribed. Housekeeping genes are being expressed in essentially every cell, while other genes are cell type or tissue specific and may be expressed at certain developmental stages\cite{strachan_humanmoleculargenetics_2011}.

% CVC

Some proteins (or other molecules such as micro RNAs) are capable of determining the rate at which a gene is transcribed (transcription rate). Such proteins are called transcription factors (TFs), and the genes they regulate are called their targets. Typically, one TF will regulate the transcription rate of many targets.

Production of a specific molecule might require multiple cascades of gene regulation. The collection of all gene regulatory interactions between transcription factors and targets is called a gene regulatory network (GRN). Studying the active parts of a cell's gene regulatory network can thus reveal which dynamic processes are taking place.

\subsection{Profiling single cells}
In order to understand a biological process, it is often quite helpful to be able to profile (i.e. observe) the biomolecular components involved therein. The single-cell "omics" technologies which we have at our fingertips today originated from the convergence of two different fields, \emph{"single-cell"} and \emph{"omics"}.

\subsubsection{Single-cell}
The earliest approaches for measuring the abundance of particular molecules in \emph{single cells} used the preferred instrument of every stereotypical biologist: the microscope.
Since it was developed by Coons et al. in 1941, immunohistochemistry (IHC) has been instrumental in visualising antigen-antibody proteins\cite{coons_immunologicalpropertiesantibody_1941}. In many multicellular organisms, antibodies and antigens serve as crucial communication tools as part of the organism's immune system. A cell can present a particular type of antigen on its cell surface, which allows a particular type of antibody to bind to it.

\mycomment{Rephrase the microscope sentence (dino joke?)}

IHC (and many other biotechnologies) visualises antigen-antibody reactions by attaching particular molecules to the antibody, such as an enzyme that catalyses a colour-producing reaction, or a fluorescent chemical compound that can re-emit light upon light excitation. Using different colours (wavelengths) allows measuring expression levels of different antibodies simultaneously. Characterising cells in a quantifiable way is labour intensive; however, since it involves acquiring an image of many cells and drawing a contour around each cell (called cell segmentation). While modern implementations of IHC improve the throughput drastically by using robots to automate the image acquisition and computer software to automate cell segmentation, the procedure is still labour intensive as the robots and computer software still needs to be kept in check.

@@

Since IHC and flow cytometry, many new technologies have been developed which allow quantifying expression levels of molecules in single cells (e.g. mass cytometry, single-cell qPCR, FISH). All of these single-cell (non-omics) technologies are limited by the number of different molecules they could measure, however; and thus required handpicking the molecules of interest before performing an experiment, making the experiment biased towards the preconceptions of the experimenter.

@@

%library(dlstats)
%library(tidyverse)
%
%cran <- c("babelwhale", "diffusionMap", "dyndimred", "dynparam", "dynutils", "GillespieSSA", "GillespieSSA2", "incgraph", "princurve", "proxyC", "qsub", "SCORPIUS", "devtools", "mlr", "ParamHelpers", "remotes", "rlang")
%bioc <- c("ClusterSignificance", "monocle", "slingshot", "splatter")
%
%dl_cran <- dlstats::cran_stats(cran)
%dl_bioc <- dlstats::bioc_stats(bioc)
%
%dls <- bind_rows(dl_cran, dl_bioc %>% select(start, end, package, downloads = Nb_of_downloads)) %>% 
%mutate(role = ifelse(package %in% c("devtools", "mlr", "ParamHelpers", "remotes", "rlang", bioc), "Contributor", "Author"))
%
%min_date <- "2019-08-01"
%num_days <- as.integer(Sys.Date() - as.Date(min_date))
%summ <- dls %>% filter("2019-08-01" <= start) %>% group_by(package) %>% summarise(downloads = sum(downloads) / num_days * 365) %>% as.data.frame

@@

    \biocpkgl{ClusterSignificance}{Clus\-ter\-Sig\-nif\-i\-cance} &  & Bioc & 803 & Assess if class clusters in dimensionality reduced data representations have a separation different from permuted data \\
    \cranpkg{devtools} & ctb & CRAN & 3'775'350 & Tools to make developing R packages easier \\
    \githubpkg{soedinglab}{merlot} & ctb & Github & \notavailable & A method for reconstructing lineage-tree topologies from scRNA-seq data \\
    \cranpkg{mlr} & ctb & CRAN & 142'605 & Machine Learning in R \\
    \biocpkg{monocle} & ctb & Bioc & 35'240 & Clustering, differential expression, and trajectory analysis for single-cell RNA-Seq \\
    \cranpkg{ParamHelpers} & ctb & CRAN & 109'408 & Helpers for Parameters in Black-Box Optimization, Tuning and Machine Learning \\
    \githubpkg{kieranrcampbell}{pseudogp} & ctb & Github & \notavailable & Probabilistic pseudotime for single-cell RNA-seq \\
    \cranpkg{Rdimtools} & ctb & CRAN & 7367 & Dimension Reduction and Estimation Methods \\
    \cranpkg{remotes} & ctb & CRAN & 3'704'594 & R package installation from remote repositories, including GitHub \\
    \cranpkg{rlang} & ctb & CRAN & 11'470'763 & Functions for base types and core R and tidyverse features \\
    \githubpkg{aertslab}{SCope} & ctb & Github & \notavailable & Visualization of large-scale and high dimensional single cell data \\
    \biocpkg{slingshot} & ctb & Bioc & 11'643 & Tools for ordering single-cell sequencing \\
    \biocpkg{splatter} & ctb & Bioc & 3741 & Simple simulation of single-cell RNA sequencing data \\
    \githubpkg{farrelja}{URD} & ctb & Github & \notavailable & URD reconstructs transcriptional trajectories underlying specification or differentiation processes in the form of a branching tree from single-cell RNAseq data \\
    \githubpkg{ManuSetty}{wishbone} & ctb & Github & \notavailable & Identify bifurcating developmental trajectories from single-cell data \\\hline
  \end{tabularx}
\end{table}

>>>>

\subsection{Cell dynamics and gene regulation}
Cells are dynamic entities that can gradually produce the molecules needed to acquire new functionality. The naturally occurring cell-to-cell variability happens at the level of gene expression. Gene expression itself can be controlled at different levels (Figx), one of which is gene regulation by transcription.
%Ref: Alberts B, Johnson A, Lewis J, Raff M, Roberts K, Walter P. An Overview of Gene Control. InMolecular Biology of the Cell. 4th edition 2002. Garland Science.

Fig. x: \textbf{Levels of controlling gene expression} can happen at the level of transcription, RNA processing (i.e. splicing), RNA transport and localization, mRNA translation, mRNA degradation and protein activity.
%From: (https://www.ncbi.nlm.nih.gov/books/NBK26885/figure/A1220/?report=objectonly) Meer inhoudelijke info: https://www.ncbi.nlm.nih.gov/books/NBK26885/ Ref: Alberts B, Johnson A, Lewis J, Raff M, Roberts K, Walter P. An Overview of Gene Control. InMolecular Biology of the Cell. 4th edition 2002. Garland Science.

According to the needs of a cell, different genes are being transcribed. Housekeeping genes are being expressed in essentially every cell, while other genes are cell type or tissue specific or may be expressed in response to developmental and environmental signals\cite{strachan_humanmoleculargenetics_2011}.

Transcription factors (TFs) modulate the rate of gene transcription by binding and recruiting the transcriptional machinery to \emph{cis}-regulatory regions (enhancers, and silencers) that are typically located in the promotor region of target genes. These bindings may result in increased or decreased gene expression. 
There are several TF families of which members share structural characteristics (e.g. zink finger, helix-loop-helix).

Many TFs are commonly present in virtually all cell types (e.g. NF-κB), while others are specific for cells and developmental stages. Typically, the same TF can regulate the rate of transcription of many target genes in different cell types, indicating that these gene regulatory networks (GRNs) are dynamic. Moreover, the production of a specific molecule might require several gene regulatory cascades. Studying the active parts of a cell's GRN can thus reveal which dynamic processes are taking place within a cell.
% Ref: Lambert SA, Jolma A, Campitelli LF, Das PK, Yin Y, Albu M, Chen X, Taipale J, Hughes TR, Weirauch MT. The human transcription factors. Cell. 2018 Feb 8;172(4):650-65.
% The kappa in NF-κB wordt niet goed omgezet in pdf doc

\subsection{Profiling single cells}
Several technologies are now available to profile (i.e. observe) biomolecular components, allowing us to gain better understanding in the biological processes that take place within a cell. The single-cell "omics" technologies originated from the convergence of two different fields, \emph{"single-cell"} and \emph{"omics"}.

\subsubsection{Single-cell}
The earliest approach for measuring the abundance of a particular molecule in \emph{single cells} is the microscope.
Since its development by Coons et al. (1941), immunohistochemistry (IHC) has been instrumental in visualising antigen-antibody proteins\cite{coons_immunologicalpropertiesantibody_1941}. In many multicellular organisms, antibodies and antigens serve as crucial communication tools as part of the organism's immune system. A cell can present a particular type of antigen on its cell surface, which allows a particular type of antibody to bind to it.

IHC (and many other biotechnologies) visualises antigen-antibody reactions by attaching particular molecules to the antibody, such as an enzyme that catalyses a colour-producing reaction, or a fluorescent chemical compound that can re-emit light upon light excitation. The use of different colours (wavelengths) allows measuring expression levels of different antibodies simultaneously. Characterising cells in a semi-quantifiable way is labour intensive, however; since it involves acquiring an image of many cells and drawing a contour around each cell (called cell segmentation). While modern implementations of IHC improve the throughput drastically by using robots to automate the image acquisition and computer software to automate cell segmentation, the procedure is still labour intensive as the robots and computer software still needs to be kept in check.

@@

Besides IHC and flow cytometry, many new technologies have been developed which allow quantifying expression levels of molecules in single cells (e.g. mass cytometry, single-cell qPCR, FISH). All of these single-cell (non-omics) technologies are limited by the number of different molecules they could measure, however. Selecting molecules of interest prior to analysis, makes the experiment biased towards the preconceptions of the experimenter.

@@

%library(dlstats)
%library(tidyverse)
%
%cran <- c("babelwhale", "diffusionMap", "dyndimred", "dynparam", "dynutils", "GillespieSSA", "GillespieSSA2", "incgraph", "lmds", "princurve", "proxyC", "qsub", "SCORPIUS", "badger", "devtools", "mlr", "ParamHelpers", "remotes", "rlang")
%bioc <- c("ClusterSignificance", "monocle", "slingshot", "splatter")
%
%dl_cran <- dlstats::cran_stats(cran)
%dl_bioc <- dlstats::bioc_stats(bioc)
%
%dls <- bind_rows(dl_cran, dl_bioc %>% select(start, end, package, downloads = Nb_of_downloads)) %>% 
%mutate(role = ifelse(package %in% c("devtools", "mlr", "ParamHelpers", "remotes", "rlang", bioc), "Contributor", "Author"))
%
%min_date <- "2019-08-01"
%num_days <- as.integer(Sys.Date() - as.Date(min_date))
%summ <- dls %>% filter("2019-08-01" <= start) %>% group_by(package) %>% summarise(downloads = sum(downloads) / num_days * 365) %>% as.data.frame

@@

    \cranpkg{badger} & ctb & CRAN & ? & Query information and generate badge for using in README and GitHub Pages \\
    \biocpkgl{ClusterSignificance}{Clus\-ter\-Sig\-nif\-i\-cance} &  & Bioc & 803 & Assess if class clusters in dimensionality reduced data representations have a separation different from permuted data \\
    \cranpkg{devtools} & ctb & CRAN & 3'775'350 & Tools to make developing R packages easier \\
    \githubpkg{soedinglab}{merlot} & ctb & Github & \notavailable & A method for reconstructing lineage-tree topologies from scRNA-seq data \\
    \cranpkg{mlr} & ctb & CRAN & 142'605 & Machine Learning in R \\
    \biocpkg{monocle} & ctb & Bioc & 35'240 & Clustering, differential expression, and trajectory analysis for single-cell RNA-Seq \\
    \cranpkg{ParamHelpers} & ctb & CRAN & 109'408 & Helpers for Parameters in Black-Box Optimization, Tuning and Machine Learning \\
    \githubpkg{kieranrcampbell}{pseudogp} & ctb & Github & \notavailable & Probabilistic pseudotime for single-cell RNA-seq \\
    \cranpkg{Rdimtools} & ctb & CRAN & 7367 & Dimension Reduction and Estimation Methods \\
    \cranpkg{remotes} & ctb & CRAN & 3'704'594 & R package installation from remote repositories, including GitHub \\
    \cranpkg{rlang} & ctb & CRAN & 11'470'763 & Functions for base types and core R and tidyverse features \\
    \githubpkg{aertslab}{SCope} & ctb & Github & \notavailable & Visualization of large-scale and high dimensional single cell data \\
    \biocpkg{slingshot} & ctb & Bioc & 11'643 & Tools for ordering single-cell sequencing \\
    \biocpkg{splatter} & ctb & Bioc & 3741 & Simple simulation of single-cell RNA sequencing data \\
    \githubpkg{farrelja}{URD} & ctb & Github & \notavailable & URD reconstructs transcriptional trajectories underlying specification or differentiation processes in the form of a branching tree from single-cell RNAseq data \\
    \githubpkg{ManuSetty}{wishbone} & ctb & Github & \notavailable & Identify bifurcating developmental trajectories from single-cell data \\\hline
  \end{tabularx}
\end{table}
